SPANG Template Library: Making SPARQL Queries FAIR

### 2020-05-23 paper.tex
v1.6

\subsection{SPARQL metadata}

@@

\subsection{SPARQL parameterization}

@@

\subsection{SPANG library API}
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

@@

\subsection{SPANG query API}

@@

\subsection{SPANG portal}
SPANG is originally developed for "SARQLing" the RDF dataset from command line. Here, we re-implemented the SPANG using Node.js and now the functionality is available through Web browser.

@@

\subsection{Template specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustash notation.

\subsection{Template engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

\subsection{Coding rules}
We defined a coding rule for SPARQL and its extension in SPANG framework. Coding rule is important for productivity of shared coding tasks and effective maintenance. It also helps to make debug process easier; weird look suggests possible bugs. 
\begin{itemize}
    \item Header lines: query can start with header lines as comment lines. Metadata can be added at the top of the query as comment lines. Multiple lines are allowed. If multiple lines are described, they are written in sequential without blank lines. After the metadata lines, a blank line comes.
    \item Prefix: prefix declaration comes in sequential lines. Each of before and after the prefix declaration lines, a blank line exists.
    \item Cases: the SPARQL keywords should be in upper case characters. variable names comes with ? followed by lower case characters.
    \item Literal: the literals are surrounded by "".
    \item Indent: indentation depth is two spaces.
    \item Spaces: a space exists between keywords and variables. A space exists before , or ; or . .
    \item Newline: a newline exists before the keyword \texttt{WHERE} and after \texttt{\{}. A new line exist after ; or . but not after ,. After the newline following ;, an indent follows. A newline exist before query modifiers such as \texttt{ORDER BY} or \texttt{LIMIT}.
\end{itemize}
We developed a reformatter of the SPARQL code as described below.

@@

\subsection{JavaScript Application}
SPANG is now implemented in JavaScript, so the SPARQL queries can be executed in JavaScript code across the internet. To avid CORS retriction, it can access CDN services such as jsDelivr~\cite{jsdelivr} as shown in Figure~\ref{fig:jsdelivr}.

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
spang.getTemplate('https://raw.githubusercontent.com/hchiba1/spang-library/master/li
brary/uniprot/count_org.rq', (template) => {
result = spang.query(template, 'https://sparql.uniprot.org', { abbr: true, format:
'tsv'}, (error, res, body) => { console.log(body); } );
})
\end{verbatim}
\end{scriptsize}
\caption{Call SARQL in a JavaScript application}
\label{fig:jsdelivr}
\end{figure}

\subsection{Command-line Client}
As an interface the the SPANG library, we constructed a command-line interface.
Originally, SPANG was developed as a command line client using Perl language~\cite{spang}.
It is re-implemented in JavaScript using Node.js as SPANG2.
SPARQL templates can be executed in command line.
The following example command line gives a parameter to the query template, and execute to the DisGeNet endpoint

\texttt{spang2 gene\_disease.rq --param disease=C0751955}

The template can be URL that returns a SPARQL query. The full usages of the command is as shown in Figure~\ref{fig:spang-command}.

@@

\end{verbatim}
\end{scriptsize}
\caption{SPANG command usage}
\label{fig:spang-command}
\end{figure}

>>>>

\subsection{SPARQL Metadata}

@@

\subsection{SPARQL Parameterization}

@@

\subsection{SPANG Library API}
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

@@

\subsection{SPANG Query API}

@@

\subsection{SPANG Server}
SPANG is originally developed for "SARQLing" the RDF dataset from command line. Here, we re-implemented the SPANG using Node.js and now the functionality is available through Web browser.

@@

\subsection{Template Specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustash notation.

\subsection{Template Engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

\subsection{Coding Rules}
We defined a coding rule for SPARQL and its extension in SPANG framework. Coding rule is important for productivity of shared coding tasks and effective maintenance. It also helps to make debug process easier; weird look suggests possible bugs. 
\begin{itemize}
    \item Header lines: query can start with header lines as comment lines. Metadata can be added at the top of the query as comment lines. Multiple lines are allowed. If multiple lines are described, they are written in sequential without blank lines. After the metadata lines, a blank line comes.
    \item Prefix: prefix declaration comes in sequential lines. Each of before and after the prefix declaration lines, a blank line exists.
    \item Cases: the SPARQL keywords should be in upper case characters. variable names comes with ? followed by lower case characters.
    \item Literal: the literals are surrounded by "".
    \item Indent: indentation depth is two spaces.
    \item Spaces: a space exists between keywords and variables. A space exists before , or ; or . .
    \item Newline: a newline exists before the keyword \texttt{WHERE} and after \texttt{\{}. A new line exist after ; or . but not after ,. After the newline following ;, an indent follows. A newline exist before query modifiers such as \texttt{ORDER BY} or \texttt{LIMIT}.
\end{itemize}
We developed a reformatter of the SPARQL code as described below.

@@

\subsection{JavaScript Application}
SPANG is now implemented in JavaScript, so the SPARQL queries can be executed in JavaScript code across the internet. To avid CORS retriction, it can access through content delivery network services, such as jsDelivr~\cite{jsdelivr}, as shown in Figure~\ref{fig:jsdelivr}.

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
spang.getTemplate('https://cdn.jsdelivr.net/gh/hchiba1/spang-library@latest/library/uniprot/count_org.rq', 
(template) => {
  result = spang.query(template, 'https://sparql.uniprot.org', { abbr: true, format:'tsv'}, 
  (error, res, body) => { console.log(body); });
});
\end{verbatim}
\end{scriptsize}
\caption{Use of SPARQL library in a JavaScript application}
\label{fig:jsdelivr}
\end{figure}

\subsection{Command-line Client}
As an interface the the SPANG library, we constructed a command-line interface.
Originally, SPANG was developed as a command line client in Perl~\cite{spang}.
It is re-implemented in JavaScript using Node.js as \texttt{SPANG2}.
SPARQL templates can be executed in command line.
The following example command line gives a parameter to the query template, and execute to the DisGeNet endpoint

\texttt{spang2 disease\_gene.rq --param disease=C0751955}

The template can be URL that returns a SPARQL query, such as \url{https://cdn.jsdelivr.net/gh/hchiba1/spang-library@latest/library/disgenet/disease_gene.rq}. The full usages of the command is as shown in Figure~\ref{fig:spang-command}.

@@

\end{verbatim}
\end{scriptsize}
\caption{Usage of the \texttt{SPANG2} command}
\label{fig:spang-command}
\end{figure}
